\documentclass[12pt,a4paper,english]{article}
\usepackage{sysload}
\hypersetup{pdftitle={System Loader - HOWTO}}

\begin{document}

\title{\Huge{}System Loader\\HOWTO}
\author{Michael Löhr and Swen Schillig}
\date{}

\maketitle
\begin{center}
\verb|$Id: howto.tex,v 1.1.1.1 2008/05/15 10:14:55 schmichr Exp $|
\end{center}
\thispagestyle{empty}
\newpage

\thispagestyle{empty}
\tableofcontents
\thispagestyle{empty}
\newpage

\sloppy


\section{Overview}
This document describes the installation and configuration of 
System Loader. All features and options of the System Loader configuration
are explained in detail. In addition, the usage of the user interface
is outlined and example configurations are provided in the appendix.


\section{Description}\label{sec:description}

\texttt{System Loader} provides a Linux system as a boot environment. This 
provides various advantages, which are explained in more detail later:
\begin{itemize}
\item no need to reinitialize after kernel update
\item more comfortable boot menu
\item network based access to boot menu
\item network based boot via ftp, http, ssh
\item possibility for centralized configuration
\item generic rescue system
\end{itemize}


Started as an extension to the \texttt{kboot} project\footnote{kboot 
homepage \url{http://kboot.sourceforge.net/}}, System Loader is now a 
stand-alone project. Like \texttt{kboot} it acts as a second stage boot 
loader by running in a minimal Linux environment started by a platform 
specific first stage boot loader. At the start a configuration file is 
read, the setup is performed according to the configuration data and finally
the boot selection menu is displayed. Based on the users selection,
a boot configuration is loaded and the system is reinitialized via
the \texttt{kexec} system call. As the result the minimal Linux system
is replaced by the system selected in the boot configuration.

\texttt{System Loader} has been optimized for a different goal than the 
original \texttt{kboot} environment. \texttt{kboot} tries to provide a compact
and minimal Linux boot system that will be built and customized for
the target machine and application. \texttt{System Loader} tries to offer
a more generic and feature rich Linux boot environment that can be
provided in a ready to use form, e.g. as a rpm-package but gives up
the goal to provide a very small boot environment.


Since \texttt{System Loader} provides a customizable Linux environment, it 
can be used in numerous ways:

\begin{enumerate}
\item \emph{As a comfortable boot loader for systems that offer only a frugal
first stage boot loader:}

\texttt{System Loader} solves the problem that
some boot loaders have to be reinitialized whenever a component of
the boot configuration changes. Examples for this type of boot loaders
are \texttt{lilo} on the i386 platform or \texttt{zipl} on the s390
platform. With \texttt{System Loader} these boot loaders will be initialized
only once to start \texttt{System Loader}. Any further changes to the
boot configuration will be handled dynamically by \texttt{System Loader}.
A potentially uncomfortable boot menu like in the case of the \texttt{zipl}
boot loader is replaced by the more convenient boot selection menu
of \texttt{System Loader}.

\item \emph{As a boot loader that can be controlled from remote via the
network:}

Usual boot managers offer their boot selection menu on the local console
of the system to be booted. Therefore it is necessary to walk to the
machine or to connect with a special terminal software if anything
different than the default configuration has to be booted. With 
\texttt{System Loader} a standard ssh interface allows remote connections 
while a system is still in its boot phase.

\item \emph{As a boot loader that supports network based boot and centralized 
boot configurations:}

Typical boot loaders have to be configured locally on the system to
be booted. If kernel parameters or the kernel itself have to be changed
on several systems an administrator has to login to all systems in
order to change the settings or install the kernel. \texttt{System Loader}'s
ability to access remote files through a network connection together
with a mechanism that can handle system specific configurations is
the basis to consolidate the boot configuration for several systems.
Boot parameters, kernels, ramdisks, etc. can be provided by a centralized
boot server. Especially in virtualization 
or cluster environments with many similar systems this allows to manage and 
change boot configurations very efficiently. The approach helps also to 
distribute new kernel versions with important bug fixes fast and keeps all 
systems booting from the same boot server in a consistent state.

\item \emph{As a rescue system that is always available in the boot sequence:}

If the boot of a Linux system fails, a Linux rescue system that allows
to fix the problem is quite useful. On a system that uses \texttt{System Loader}
as its boot manager there is already a rescue system installed. From
the shell environment of \texttt{System Loader} the problem that prevents
the system from booting can be analyzed and fixed. As soon as the
problem is solved the system can be booted into normal operation again.
Special tools and requirements for a customized rescue system can
be easily integrated into the \texttt{System Loader} ramdisk via the 
\texttt{sysload\_admin} tool.
\end{enumerate}



\section{Building and Installation}
The README file provided with \texttt{System Loader} describes in 
detail, how \texttt{System Loader} can be built. 


\subsection{Manual Installation}
\texttt{System Loader} can be easily installed by copying a 
\texttt{System Loader} RAM disk from another system. Depending on how 
the first stage boot loader is configured, a special kernel and a 
\texttt{System Loader} configuration file also need to be copied. It 
is however possible to use the kernel already existing and to utilize 
a central configuration file accessible through network as explained 
later in this document.

Sections \ref{sub:menu.lst-for-grub} and \ref{sub:zipl.conf} show 
example configurations for \texttt{grub} and \texttt{zipl}.
In some cases, the first stage boot loader (e.g. \texttt{lilo} or 
\texttt{zipl}) needs to be reinitialized after the configuration file 
has been adjusted. 

This installation type will provide an environment which allows using 
\texttt{System Loader} as a second level boot loader. However, it will 
not provide any means for modifying the RAM disk. This can be achieved 
with one of the following installation methods.


\subsection{Installing from source}
It is also easy to install \texttt{System Loader} from source. All 
Makefiles provide the targets \texttt{install} and \texttt{uninstall}.
\texttt{install} will install all files necessary for creating and 
modifying the \texttt{System Loader} RAM disk. \texttt{uninstall} will 
remove all these files from the system.

However, \texttt{make install} will not initialize the \texttt{System Loader}
RAM disk. After installing, a \texttt{sysload\_admin.conf} has to be copied 
to \texttt{/etc} and adjusted to the system and the \texttt{System Loader} 
administration tool \texttt{sysload\_admin} needs to run to create the 
required RAM disk. Also, the configuration file of the first stage boot 
loader needs to be adjusted and the boot loader needs to be reinitialized.


\subsection{Installing a RPM}
It is also possible to install \texttt{System Loader} using a RPM 
package. A RPM package can be build using the spec file which is provided 
along with \texttt{System Loader}'s source. A detailed instruction on how 
to build the package can be found in the README file.
 
RPM packages can be installed with the \texttt{rpm} tool on systems using 
RPM for package management and with the \texttt{alien} tool an many other 
systems, using e.g. Debian packages.



\section{Boot Methods}
Depending on the hardware platform and the available disk types the
System Loader supports several boot methods. On the i386 and s390
platform the system can be booted from a kernel image file. An initial
ramdisk and a kernel command line or parmfile can be specified together
with this kernel file. On the s390 platform \texttt{*.ins} files
or the boot map information written by \texttt{zipl} can be used in
addition.

In any case the location of kernel, ramdisk, parmfile, insfile or
boot map will be defined as an URI. Which URI scheme has to be used
depends on the system platform and available disk. The \texttt{block}
URI is typically used on the i386 platform. The \texttt{dasd} and
\texttt{zfcp} URI schemes are specific for the s390 platform. On every
platform the \texttt{ftp} , \texttt{http} and \texttt{scp} URI schemes
can be used to boot from the network.

All boot methods and URI schemes will be described in detail in the
following chapter.



\section{Elements of the Config File}\label{sec:Elements-of-the}
The config file for \texttt{System Loader}
consists of some global definitions which are followed by one or more
boot entries. External files are always referenced by an URI. A detailed
description of supported URIs will be given in section \ref{sub:URIs}.


\subsection{Global Definitions}
Preceding the boot entries the System Loader configuration file contains
a number of definitions that are valid for the whole configuration
file.


\subsubsection{Comments}
Comments are allowed in the global definitions part of the configuration
file and in the boot entries. A comment starts with \texttt{\#} and
ends at the end of the line. Tabs or spaces in front of the \# are
possible but comments at the end of a line that contains a definition
are not allowed.

Example:
\begin{verbatim}
# this is a comment
        # this is also a comment

<definition> # this is not allowed!
\end{verbatim}


\subsubsection{\texttt{default}}\label{sub:default}
The \texttt{default} statement references the boot entry that will
be selected automatically after the timeout. Valid references are
the label of a boot entry or a number. Implicit numbering of the boot
entries starts with 0. If the default statement is not used, entry
number 0 will be used as the default.

Example:
\begin{verbatim}
default linux2
default 1
\end{verbatim}


\subsubsection{\texttt{timeout}}
The \texttt{timeout} statement specifies the time in seconds until
the default boot entry will be started automatically. If the statement
is not used or the timeout is set to 0 no timeout will occur.

Example:
\begin{verbatim}
timeout 15
\end{verbatim}


\subsubsection{\texttt{password}}\label{sub:password}
The \texttt{password} statement defines the password that has to be
entered if a locked boot entry (see section \ref{sub:lock}) has been
selected.

Example:
\begin{verbatim}
password topsecret
\end{verbatim}


\subsubsection{\texttt{userinterface}}
The \texttt{userinterface} statement specifies which userinterface
module should be started to display the boot selection menu. For every
userinterface command a userinterface process will be started. This
allows to start several instances of the same userinterface module
listening on different input devices or ports. Additional options
may be specified depending on the user interface module. Currently
\texttt{linemode} and ssh are available as user interface modules.

Syntax:
\begin{verbatim}
userinterface <ui_module_name> [MODULE_OPTIONS]
\end{verbatim}


\subsubsection{\texttt{userinterface linemode}}
For the linemode interface it is required to specify the device on
which the boot selection menu will be displayed. Typical values for
the device parameter are \texttt{/dev/console}, \texttt{/dev/tty}
or \texttt{/dev/tty1}.

Syntax:
\begin{verbatim}
userinterface linemode <device>
\end{verbatim}

Example:
\begin{verbatim}
userinterface linemode /dev/tty1
\end{verbatim}


\subsubsection{\texttt{userinterface ssh}}
The \texttt{userinterface ssh} statement will start a process that
is listening for incoming ssh connections in the background. For every
incoming connection that is successfully established a new user interface
instance will be started. The type of interface to be started has
to be specified as the first module option. Currently only linemode
can be specified here but for the future additional user interface
types (e.g. ncurses) may become available. Additional module options
may be used to specify the port on which the ssh connection can be
established and to load RSA or DSS keys from a location defined by
an URI scheme.

Connecting to System Loader via ssh will require a userid and password. 
During the creation of the System Loader ramdisk a user \texttt{sysload} 
with password \texttt{sysload} will be created automatically. The 
password may be changed with the \texttt{sysload\_admin} tool. 

Syntax:
\begin{verbatim}
userinterface ssh <ui_module_name> [port=<portnumber>] \
                  [rsa_key=<key_uri>] [dss_key=<key_uri>]
\end{verbatim}

Example:
\begin{verbatim}
userinterface ssh linemode port=2222
\end{verbatim}


\subsubsection{\texttt{include}}
The \texttt{include} statement can be used anywhere in the system
loader configuration file. The content of the file referenced by the
URI will be included at the position of the \texttt{include} statement.
Nested includes are allowed but limited to twenty levels. \texttt{include}
statements that appear inside an inactive system section will not
be applied.

Syntax:
\begin{verbatim}
include URI
\end{verbatim}

Example:
\begin{verbatim}
include dasd://(0.0.5c5e,1)/boot/extra_menu.conf
\end{verbatim}


\subsubsection{\texttt{exec}}
The \texttt{exec} statement can be used to execute additional commands
while the System Loader configuration file is evaluated. The specified
command is executed immediately. The evaluation of the configuration
file continues as soon as the specified command returns. \texttt{exec}
statements that appear inside an inactive system section will not
execute the specified command.

Syntax:
\begin{verbatim}
exec <command>
\end{verbatim}

Example:
\begin{verbatim}
exec mdadm --assemble /dev/md0 /dev/sda2 /dev/sdb2
exec vgscan
exec vgchange -a y
\end{verbatim}



\subsection{Setup Commands}
This group of commands allows to setup additional devices in the minimal
Linux environment of the System Loader. It is possible to load additional
kernel modules, to specify network settings and to enable devices
that are not enabled automatically. Each \texttt{setup} command is
executed immediately. As parameters are identified by keywords they
are not required to appear in any particular order. An additional
short form allows to use \texttt{setup} commands on the kernel command
line.

Syntax:
\begin{verbatim}
setup <setup_item> { 
  <paramlist> 
}
\end{verbatim}


\subsubsection{\texttt{setup module}}
The \texttt{setup module} command allows to load additional kernel
modules. Module dependencies are resolved automatically. The module
to be loaded has to be present on the System Loader ramdisk. Modules
can be added to an existing ramdisk via the sysload\_admin tool.

Syntax:
\begin{verbatim}
setup module {
  name          <name>
  param         <param>
  kernelversion <kernelversion>
}
\end{verbatim}

Short form:
\begin{verbatim}
mod(<name>[,<param>[,<kernelversion>]])
\end{verbatim}

Example:
\begin{verbatim}
setup module {
  name qeth
}
\end{verbatim}


\subsubsection{\texttt{setup network}}
The \texttt{setup network} command allows to specify network settings
and to initialize a network device with these settings. Static setup
and dhcp are supported. Parameters for static network setup can also
be used in dhcp mode and are used as fallback if dhcp fails. A correct
network setup is required to use network based URI schemes like ftp,
http and scp. A \texttt{setup module} command may be required to load
the network device driver before the \texttt{setup network} command
can be executed successfully. The \texttt{sysload\_admin} tool can be
used to copy the required driver modules into the ramdisk of the system
loader.

Syntax:
\begin{verbatim}
setup network {
  interface  <interface>
  mode       <dhcp_or_static>
  address    <address>
  mask       <mask>
  gateway    <gateway>
  nameserver <nameserver>
}
\end{verbatim}

Short form:
\begin{verbatim}
static(<interface>[,<address>[,<mask>[,<gateway>[,<nameserver>]]]])
dhcp(<interface>[,<address>[,<mask>[,<gateway>[,<nameserver>]]]])
\end{verbatim}

Example:
\begin{verbatim}
setup network {
  interface  eth0
  mode       dhcp
  address    9.155.23.65
  mask       255.255.255.128
  gateway    9.155.23.1
  nameserver 9.64.163.21
}
\end{verbatim}


\subsubsection{\texttt{setup qeth}}
On the s390 platform the \texttt{setup qeth} command can be used to
enable a qeth Ethernet device with the given busids.

Syntax:
\begin{verbatim}
setup qeth {
  busid  <busid>
  busid  <busid>
  busid  <busid>
}
\end{verbatim}

Short form:
\begin{verbatim}
qeth(<busid>,<busid>,<busid>)
\end{verbatim}

Example:
\begin{verbatim}
setup qeth {
  busid 0.0.f5de
  busid 0.0.f5df
  busid 0.0.f5e0
}
\end{verbatim}


\subsubsection{\texttt{setup dasd}}
On the s390 platform the \texttt{setup dasd} command can be used to
enable a dasd disk device with the given busid.

Syntax:
\begin{verbatim}
setup dasd {
  busid  <busid>
}
\end{verbatim}

Short form:
\begin{verbatim}
dasd(<busid>)
\end{verbatim}

Example:
\begin{verbatim}
setup dasd {
  busid 0.0.5c60
}
\end{verbatim}


\subsubsection{\texttt{setup zfcp}}
On the s390 platform the \texttt{setup zfcp} command can be used to
enable a zfcp disk device.

Syntax:
\begin{verbatim}
setup zfcp {
  busid  <busid>
  wwpn   <wwpn>
  lun    <lun>
}
\end{verbatim}

Short form:
\begin{verbatim}
zfcp(<busid>,<wwpn>,<lun>)
\end{verbatim}

Example:
\begin{verbatim}
setup zfcp {
  busid 0.0.54ae
  wwpn  0x5005076300cb93cb
  lun   0x512e000000000000
}
\end{verbatim}



\subsection{System Dependent Sections}
Based on the network capabilities of System Loader it is possible
to provide kernels, ramdisks and kernel parameters on a centralized
bootserver. System dependent sections in the System Loader configuration
file provide a mechanism that is especially useful for shared system
loader configuration files. A system dependent section starts with
a test, that determines if the configuration part is active on a specific
system. Depending on the result of this test the following section
is used or ignored. The identification of the system can be done via
the MAC address of a network adapter, the UUID of the system, the
name of the virtual machine or the name of the logical partition of
the host system. A system dependent section is active if at least
one of the system identifiers matches. If a system identifier is not
available on the specific system it will never match. System dependent
sections can be nested in order to specify settings that are common
to a number of systems and to be able to change only some details
for the specific system.

Syntax:

\begin{verbatim}
system <systemidlist> {
  <definitionlist>
}
\end{verbatim}

Examples:
\begin{verbatim}
system mac(00:10:C6:DE:12:6C) 
       uuid(C7CCA781-2DD5-11C6-93BC-AD439DC0988B)
{
  root block://(/dev/hda5,ext3)/boot/
}

system not(vmguest(linux41)) {
  include ftp://sysload:sysload@53v15g41.ibm.com/boot/standard\_entries.conf
}

system vmguest(linux40,g53lp15)
       vmguest(linux41,g53lp15)
{
  setup network {
    mode static

    vmguest(linux40) {
      address   9.152.26.120
    }

    vmguest(linux41) {
      address   9.152.26.121
    }

    mask    255.255.252.0
    gateway 9.152.24.1
    nameserver 9.152.120.241
    interface eth0
  }
}
\end{verbatim}


\subsubsection{\texttt{mac}}
The mac systemidentifier can be used on nearly every system that has
a network adapter. The mac statement matches if the system has a network
adapter with this mac address. It may be necessary to load a kernel
module and to enable the network device before the MAC of the system
can be determined. Therefore it is not the preferred method to identify
a system via its mac. Nevertheless it is the only available method
for older machines on the i386 platform.


\subsubsection{\texttt{uuid}}
On the i386 platform recent BIOS versions provide a UUID that allows
to identify the specific system unambiguously. The \texttt{uuid} command
matches when the given UUID matches the UUID of the system. If available
this is the preferred method to identify a specific system on the
i386 platform.


\subsubsection{\texttt{lpar}}
The concept of logical partitions allows to split the hardware of
a computer system into several independent parts. Each part or logical
partition can boot its own operating system. The \texttt{lpar} system
identifier allows to identify the logical partition on which the system
loader configuration file is currently used and matches when the given
name matches the name of the current logical partition. In the system
loader it is implemented for the s390 platform. In the future it could
be extended to other platforms that offer logical partitions. The
name of the lpar is read from \texttt{/proc/sysinfo.}

Syntax:
\begin{verbatim}
lpar(lparname)
\end{verbatim}


\subsubsection{\texttt{vmguest}}
The concept of virtual machines is an approach that allows to run
several instances of operating systems under the control of a virtualization
software. The \texttt{vmguest} system identifier allows to identify
the virtual machine on which a System Loader configuration file is
currently used. It matches when the given name matches the name of
the virtual machine. An extended form allows to specify the name of
a logical partition in addition and matches when the name of the virtual
machine and the name of the logical partition are identical to the
given parameter values. The \texttt{vmguest} system identifier is
implemented for VM on the the s390 platform. In the future it could
be extended to other environments that offer virtual machines. The
names of the vmguest and of the lpar are read from \texttt{/proc/sysinfo.}

Syntax:
\begin{verbatim}
vmguest(vmguestname)
\end{verbatim}

or
\begin{verbatim}
vmguest(vmguestname, lparname)
\end{verbatim}



\subsection{Boot Entry Definitions}
The global definitions are followed by a section of boot entries which
describe the available boot configurations. A boot entry starts with
the line

\begin{verbatim}
boot_entry {
\end{verbatim}
and ends with a closing bracket
\begin{verbatim}
}
\end{verbatim}

Every boot entry must contain exactly one statement that triggers
a boot action. These boot action statements are \texttt{kernel}, \texttt{insfile},
\texttt{bootmap}, \texttt{halt}, \texttt{reboot} and \texttt{shell}.
Other statements can be used optionally. Some of them will specify
labels or variables to improve the handling of the configuration file,
some will influence the behaviour of the boot menu and some will pass
additional information to the booted system. All possible statements
inside the boot entry are described in the following sections.

The length of the complete boot entry can not exceed the maximum string 
length which is set to 512 characters by default. There is a global 
parameter which sets the string length in \texttt{config.h}.

\subsubsection{\texttt{title}}
The \texttt{title} statement defines a text that will be displayed
in the boot selection menu and should give some meaningful description
of the boot configuration. This statement is mandatory for all boot
entries.

Example:
\begin{verbatim}
title Debian GNU/Linux, latest kernel
\end{verbatim}


\subsubsection{\texttt{label}}
The \texttt{label} statement allows to assign a symbolic label to
a boot entry. This label can be used in the default statement to reference
the default boot entry (see section \ref{sub:default}). If no label
definition is present the boot entry can still be referenced numerically
by its position in the configuration file.

Example:
\begin{verbatim}
label linux3
\end{verbatim}


\subsubsection{\texttt{root}}
The \texttt{root} statement defines an URI path prefix and will be
prepended to all URIs%
\footnote{The root definition will not be applied to \texttt{bootmap} URIs because
a path definition makes no sense for a structure that is not part
of any filesystem.%
} specified in the same boot entry. Typically it is used to specify
a common path for kernel, initrd and parmfile. Note that the path
prefix and the rest of the URI will be concatenated as they are specified.
There is no automatic insertion of a \texttt{'/'} character and no
syntax checking.

Example:
\begin{verbatim}
root dasd://(0.0.5c5e,1)/boot/
\end{verbatim}


\subsubsection{\texttt{kernel}}
The \texttt{kernel} statement specifies the kernel file that will
be booted if this boot entry is selected. All supported URI formats
are allowed to specify the location of the kernel file. If a root
statement is given in the same boot entry, it will be prepended to
the specified kernel path.

Example:
\begin{verbatim}
kernel vmlinuz
\end{verbatim}


\subsubsection{\texttt{initrd}}
The \texttt{initrd} statement specifies the initial ramdisk that will
be used if this boot entry is selected. All supported URI formats
are allowed to specify the location of the ramdisk file. If a root
statement is given in the same boot entry, it will be prepended to
the specified ramdisk path.

Example:
\begin{verbatim}
initrd initrd.img
\end{verbatim}

\subsubsection{\texttt{cmdline}}
The cmdline statement specifies the kernel commandline that is used
to start the kernel if this boot entry is selected.

Example:
\begin{verbatim}
cmdline ro root=/dev/ram0 ramdisk_size=100000
\end{verbatim}


\subsubsection{\texttt{parmfile}}
As an alternative to the cmdline statement the \texttt{parmfile} statement
can be used to specify the file that will be used as the kernel command
line if this boot entry is selected. All supported URI formats are
allowed to specify the location of the parmfile. If a root statement
is given in the same boot entry, it will be prepended to the specified
parmfile path. If parmfile and cmdline are specified at the same time
they will be concatenated as \texttt{parmfile} + \texttt{' '} + \texttt{cmdline}.

Example:
\begin{verbatim}
parmfile parmfile.txt
\end{verbatim}


\subsubsection{\texttt{lock}}\label{sub:lock}
The \texttt{lock} command allows to have password protected boot entries.
If a locked boot entry is selected the user has to enter the password
specified in the \texttt{password} statement (see section \ref{sub:password})
to execute this boot selection entry.

Example:
\begin{verbatim}
lock
\end{verbatim}


\subsubsection{\texttt{pause}}
The \texttt{pause} statement displays a message and waits for user
input before the boot entry will be started. This can be used to ask
the user to prepare the system for booting (e.g. by inserting a boot
CD to the CD drive).

Example:
\begin{verbatim}
pause Please insert your boot floppy!
\end{verbatim}


\subsubsection{\texttt{insfile}}
The \texttt{insfile} statement can be used as an alternative method
to specify a boot configuration. An \texttt{*.ins} file
contains the definitions of a kernel, initial ramdisk and parmfile,
therefore it makes no sense to specify these components . The \texttt{insfile}
statement is only available on the s390 platform and is provided with
several Linux distributions for this platform.

Example:
\begin{verbatim}
insfile dasd://(0.0.5c5e,1)/usr/local/insfile_test/redhat/generic.ins
\end{verbatim}


\subsubsection{\texttt{bootmap}}\label{sub:bootmap}
The \texttt{bootmap} command can be used as an alternative method
to specify a boot configuration. It is only available on the s390
platform and boots the system using the boot information from the
boot map of the specified disk. This boot information is written by
the first stage boot loader \texttt{zipl}. Only bootmaps in the format 
created by \texttt{zipl} version 1.2 or newer are supported. The 
bootmap command uses a modified URI format that is described in 
section \ref{sub:dasd-URI-for} and \ref{sub:zfcp-URI-for}.

Example:
\begin{verbatim}
bootmap dasd://(0.0.5e2a,0)
\end{verbatim}


\subsubsection{\texttt{halt}}
The \texttt{halt} statement can be used instead of a real boot selection.
It will halt the Linux environment of the System Loader.


\subsubsection{\texttt{reboot}}
The \texttt{reboot} statement can be used instead of a real boot selection.
This statement will reboot the system. If a first stage bootloader
is installed the system will be restarted via this bootloader.


\subsubsection{\texttt{shell}\label{sub:shell}}
The \texttt{shell} statement can be used instead of a real boot selection.
If the resulting boot entry is selected, the user will get a shell
prompt. Leaving the shell prompt via exit will redisplay the boot
selection menu.



\subsection{URI Definitions}\label{sub:URIs}
URIs provide a generic mechanism to describe the
location of files. They are used by System Loader to locate boot
and configuration files. The following sections describe the supported
URI schemes in detail.


\subsubsection{\texttt{block}}
The \texttt{block} URI can be used to reference a file on any block
device containing a supported filesystem. This is the typical method
to access the boot files on the i386 platform and on other non s390
platforms. If no filesystem type is specified, the filesystem type
will be auto detected.

Syntax:
\begin{verbatim}
block://(<device node>[,<filesystem type>])/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
block://(/dev/hda5,ext3)/boot/vmlinuz
\end{verbatim}


\subsubsection{\texttt{ftp}}\label{sub:URI_ftp}
The \texttt{ftp} URI can be used to reference a file on a ftp-server.
This URI type is available on all platforms but requires a network
connection. If the remote server requires a UID and a password, it
has to be provided the usual way (see the example below). In case
the UID contains the ``@'' , as it happens in e-mail addresses,
it has to be replaced with ``\%40'' to avoid any misunderstanding
with the separator between the UID:PW combination and the host-name.

Syntax:
\begin{verbatim}
ftp://[UID[:PW]@]hostname[:PORT]/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
ftp://my_name%40ibm.com:my_secret@ftp.server.com/boot/vmlinuz
\end{verbatim}


\subsubsection{\texttt{http}}\label{sub:URI_http}
The \texttt{http} URI can be used to reference a file on a http-server.
This URI type is available on all platforms but requires a network
connection. If the remote server requires a UID and a password, it
has to be provided like for the \ref{sub:URI_ftp}ftp URI (see example
below). In case the UID contains the ``@'' , as it happens in
e-mail addresses, it has to be replaced with ``\%40'' to avoid
any misunderstanding with the separator between the UID:PW combination
and the host-name.

Syntax:
\begin{verbatim}
http://[UID[:PW]@]hostname[:PORT]/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
http://my_name%40ibm.com:my_secret@http.server.com/boot/vmlinuz
\end{verbatim}


\subsubsection{\texttt{scp}}
The \texttt{scp} URI can be used to reference a file on a server which
is accessible via the SSH protocol. As with the \ref{sub:URI_ftp}
ftp- and the \ref{sub:URI_http} http-URI the scp-URI is available
on all platforms but requires a network connection. The encoding of
a UID and a password is identical to the above described protocols
(see example below). In case the UID contains the ``@'' , as it
happens in e-mail addresses, it has to be replaced with ``\%40''
to avoid any misunderstanding with the separator between the UID:PW
combination and the host-name. Beside the UID:PW combination, encoded
in the URI, the public key authentication is supported as well. Therefor
the local public key has to be part of the servers \texttt{authorized\_keys}
file and the public key authentication has to be enabled on the server.

Syntax:
\begin{verbatim}
scp://[UID[:PW]@]hostname[:PORT]/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
scp://my_name%40ibm.com:my_secret@ssh.server.com/boot/vmlinuz
\end{verbatim}


\subsubsection{\texttt{file}}
The \texttt{file} URI can be used to reference files on an already
mounted filesystem. This type of URI is available on all platforms.
In an unmodified System Loader boot environment it can only be used
to reference files on the initial ramdisk. If the System Loader is
running as a program on an already booted Linux system it can be used
instead of the other URIs.

Example:
\begin{verbatim}
file:///boot/vmlinuz
\end{verbatim}


\subsubsection{\texttt{dasd}}
The \texttt{dasd} URI can be used to reference files on zSeries ESCON/FICON
attached storage. This type of URI is only available on the s390 platform.
If no filesystem type is specified, the filesystem type will be auto detected.

Syntax:
\begin{verbatim}
dasd://(<bus id>,[<partition>[,<filesystem type>]])/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
dasd://(0.0.5c5e,1)/usr/local/insfile_test/SuSE/suse.ins
\end{verbatim}


\subsubsection{\texttt{zfcp}}
The zfcp URI can be used to reference files on s390 (zSeries) FCP
attached storage. This type of URI is only available on the s390 platform.
If no filesystem type is specified, the filesystem type will be auto detected.

Syntax:
\begin{verbatim}
zfcp://(<bus id>,<WWPN>,<LUN>,[<partition>[,<filesystem type>]])/<path to file>
\end{verbatim}

Example:
\begin{verbatim}
zfcp://(0.0.54e0,0x5005076303000104,0x4011400500000000,1)/boot/initrd.img
\end{verbatim}


\subsubsection{\texttt{dasd} URI for the bootmap command}\label{sub:dasd-URI-for}
Because the bootmap information is not stored inside a file system
the bootmap command (see section \ref{sub:bootmap}) uses a reduced
form of the dasd URI.

Syntax:
\begin{verbatim}
dasd://(<bus id>[,<program number>])
\end{verbatim}

Example:
\begin{verbatim}
dasd://(0.0.5e89,1)
\end{verbatim}


\subsubsection{\texttt{zfcp} URI for the bootmap command}\label{sub:zfcp-URI-for}
Because the bootmap information is not stored inside a file system
the bootmap command (see section \ref{sub:bootmap}) uses a reduced
form of the zfcp URI.

Syntax:
\begin{verbatim}
zfcp://(<bus id>,<WWPN>,<LUN>[,<program number>])
\end{verbatim}

Example:
\begin{verbatim}
zfcp://(0.0.04ae,0x500507630e01fca2,0x4010404500000000,2)
\end{verbatim}



\section{Using the Command Line Interface}
The command line interface is the basic user interface of System
Loader. It is designed to run on all platforms and without any special
requirements regarding the capabilities of the user interface device.
Therefore it will be usable on line mode interfaces like the 3270
terminal on the s390 platform or via a serial line on typical open
systems platforms.

The command line interface displays the title information from all
boot entries found in the System Loader configuration file. The default
entry is marked by an arrow symbol \texttt{->}, locked entries are
displayed in brackets \texttt{{[}]}.

\begin{verbatim}
System Loader user interface is starting. 
Configuration file source: file:///boot/sysload.conf

Welcome to System Loader!

The following boot options are available:
-> 1    latest kernel from DASD
   2    latest kernel from EVMS (matching system)
   3    rescue kernel from DASD
   4    System Reboot
   5    System Halt
   6    start a shell!
   d<n> Display boot parameters of the selected entry
   m<n> Modify and boot selected entry
   i    Enter boot parameters interactively

Please enter your selection:
\end{verbatim}

The selection of a specific entry is done by entering its number.
If the menu is not completely visible it can be redisplayed by entering
an empty input. If a timeout occurs while one or more user interfaces
are waiting for input, all user interfaces will be terminated and
the default entry will be executed. If a timeout is defined it will
be displayed by the user interface. In addition to the selection of
predefined menu entries the user interface allows to display existing
boot entries, to modify them or to enter new entries from scratch.

Entering the boot entry number with prefix \texttt{d} will display
the associated boot entry like shown in the following example. 

\begin{verbatim}
d2

*** BOOT ENTRY PRINTOUT ***
title     latest kernel from EVMS (matching system)
label     evms
root      block://(/dev/evms/evmsvol)/boot/
kernel    vmlinuz
cmdline   dasd=5c60-5c61 root=/dev/dasda1 ro noinitrd selinux=0
action    KERNEL_BOOT
*** Press RETURN to continue ***
\end{verbatim}

With prefix \texttt{m} the boot entry will first be displayed completely
and will then be offered line by line to allow the input of modified
lines. Empty user input will leave the respective line unchanged.
The modified boot entry will be displayed again and can be booted
finally.

\begin{verbatim}
m2

*** BOOT ENTRY PRINTOUT ***
title     latest kernel from EVMS (matching system)
label     evms
root      block://(/dev/evms/evmsvol)/boot/
kernel    vmlinuz
cmdline   dasd=5c60-5c61 root=/dev/dasda1 ro noinitrd selinux=0
action    KERNEL_BOOT
 
*** Please enter the new values for each line ***
*** or hit ENTER to leave it unchanged.       ***
root
kernel    vmlinuz.rescue
initrd
cmdline
parmfile
 
*** BOOT ENTRY PRINTOUT ***
root      block://(/dev/evms/evmsvol)/boot/
kernel    vmlinuz.rescue
cmdline   dasd=5c60-5c61 root=/dev/dasda1 ro noinitrd selinux=0
action    KERNEL_BOOT

*** Press 'B' to boot this entry or any other key to cancel ***
\end{verbatim}

The \texttt{i} command allows to enter a boot entry interactively
from scratch in cases where no similar entry is already present in
the boot selection menu.

\begin{verbatim}
i

*** Please enter the values for manual boot ***
ACTION   => 1->KERNEL_BOOT 2->INSFILE_BOOT 3->BOOTMAP_BOOT
action    1
root      block://(/dev/evms/evmsvol)/boot/
kernel    vmlinuz.rescue
initrd
cmdline   dasd=5c60-5c61 root=/dev/dasda1 ro noinitrd selinux=0
parmfile
 
*** BOOT ENTRY PRINTOUT ***
root      block://(/dev/evms/evmsvol)/boot/
kernel    vmlinuz.rescue
cmdline   dasd=5c60-5c61 root=/dev/dasda1 ro noinitrd selinux=0
action    KERNEL_BOOT
 
*** Press 'B' to boot this entry or any other key to cancel ***
\end{verbatim}



\section{The \texttt{sysload\_admin} Tool}\label{sec:The-sysload_admin-Tool}
The administration tool was developed to create or modify a ramfs
to suite all system and user requirements while running the minimal
system contained within the ramfs. The functionality includes the
support to add any executable, library or module to the ramfs. In
addition all dependencies are detected and resolved by adding the
dependent files to the ramfs.

E.g. if an added executable is dynamically linked and has therefor
certain library dependencies, this will be detected and the required
libraries will be copied to the ramfs. 

The same applies for library (e.g. library\_a requires library\_b)
and module (mod\_a requires mod\_b) dependencies. If modules are added
to the ramfs the required modules.dep file is generated in the appropriate
location.

In addition the tool can be used to change the super-users password
within the ramfs. This functionality is primarily used when the ramfs
is supporting remote-access while running the minimal system.

The individual settings of the admin tool are configured through a
configuration file which can be specified by the '-c <file>' command-line
parameter. The default is /etc/sysload\_admin.conf.

If System Loader was newly installed, this configuration file is effectively 
empty and sysload\_admin will show errors. Please refer to the system 
specific example configurations provided by System Loader and adjust them 
as needed.

In addition the sysload\_admin tool supports the following options and
commands.

\begin{verbatim}
Commands:
  local    create a ramfs based on local source files
  passwd   change password for super-user ID in -m <ramfs>
  add      add executables, libraries and/or modules to the ramfs

Options:
  -h, --help              this help text
  -V, --version           print version information
  -c, --config <file>     configuration file name (default /etc/sysload_admin.conf)
  -m, --master <file>     master ramfs superseding other ramfs' content.
  -o, --output <file>     file name for the resulting ramfs image
  -k, --kernel <file>     kernel file name
  -l, --lib <file>        add DLL(s) <file,...> to ramfs (full qualified path)
  -d, --module <module>   add module(s) <module,...> to ramfs (e.g. qeth)
  -x, --executable <file> add executable <file,...> to ramfs (e.g. losetup)
  -v, --verbose           be verbose
  -q, --quite             be quite
\end{verbatim}

Notice: if one or more of the command-line switches \texttt{-l, -m,
-x} are used, none of the configured file values are used for libraries,
modules and executables. This means that either the configuration
file or the command line is used to reference the files which are
to be added to the ramfs.

To describe the admin tool's functionality best, verify the following
examples:

\begin{enumerate}
\item generate a ramfs containing the defaults specified in the configuration
file. 
\begin{verbatim}
sysload_admin local
\end{verbatim}

\item add the executables ldd and /home/frank/my\_exec to the ramfs /boot/my\_initrd
and store the resulting ramfs in the default file. 
\begin{verbatim}
sysload_admin -m /boot/my_initrd -x ldd,/home/frank/my_exec add
\end{verbatim}

\item add all executables, libraries and modules specified in the configuration
file /home/frank/my.conf and store the resulting ramfs in /home/frank/my\_initrd.new.
\begin{verbatim}
sysload_admin -f /home/frank/my.conf -o /home/frank/my_initrd.new add
\end{verbatim}

\item change the super-users password in the ramfs /boot/my\_initrd and
store the resulting ramfs in the same file.
\begin{verbatim}
sysload_admin -m /boot/my_initrd -o /boot/my_initrd passwd
\end{verbatim}
\end{enumerate}



\section{Using Online Documentation}
Man pages are available for the \texttt{sysload} executable and for
the System Loader configuration file \texttt{sysload.conf}. These man
pages are meant as a short overview and refer to this specification
and to the System Loader design document for a detailed description
of this software.

There are also man pages for the System Loader admin tool 
\texttt{sysload\_admin} and its configuration file 
\texttt{sysload\_admin.conf}.



\section{Extending the System Loader}
The System Loader is designed to be extensible. The extension can
be done without changing the code of the existing components. User
provided loader modules can provide support for additional URI schemes
(e.g. nfs, smb). Additional user interface modules can provide more
comfortable user interfaces (e.g. graphical or web-based). Interface
definitions are given in the design documents.



\section{Dependencies}
It is possible to use the System Loader within any 
Linux runtime environment. The main dependencies are
the availability of the kexec system call in the kernel and the kexec
tool. In addition the current implementation depends on several basic
Linux tools and programs like \texttt{udev}, \texttt{sh}, \texttt{awk},
\texttt{mount}, \texttt{umount}, \texttt{mkdir}, \texttt{rmdir}, \texttt{usleep},
etc.



\section{Performance Considerations}
System Loader may slow down the overall boot process because two
Linux kernels must be loaded until the system is running in its final
configuration. It is expected that System Loader will be primarily
used in situations where the flexibility of the boot process as 
described in section \ref{sec:description} is most important.




\appendix

\section{Examples}

\subsection{\texttt{menu.lst} to start System Loader via \texttt{grub}}
This example shows how System Loader can be booted on the i386
platform using grub as the first stage boot loader. Relevant for 
System Loader is the \texttt{sysload=} parameter in the kernel command
line which references the System Loader configuration file. The 
\texttt{quiet} parameter minimizes the screen output while the 
System Loader environment is started.

\label{sub:menu.lst-for-grub}
{\scriptsize \verbatiminput{menu.lst}}


\subsection{Simple \texttt{sysload.conf} for i386}
The following example shows a simple System Loader configuration for
the i386 platform. Two userinterface instances are started to listen
on two different devices simultaneously.

\label{i386conf}
{\scriptsize \verbatiminput{sysload_i386.conf}}


\subsection{An Example Using Advanced Configuration Features}
The following example shows how the same System Loader configuration
files can be used on two different systems. The platform for the examples
is s390 which gets the boot configuration of the first stage boot
loader in the file \texttt{zipl.conf}. Both systems share the same 
configuration files.


\subsubsection{zipl.conf on System 1\label{sub:zipl.conf}}
System 1 is a virtual machine called \texttt{linux41}. The configuration
for the first stage bootloader \texttt{zipl} references a file 
\texttt{parmfile\_linux41.sysload}
which contains the kernel commandline that is used to boot the initial
System Loader linux environment.

{\scriptsize \verbatiminput{zipl1.conf}}

The kernel command line in \texttt{parmfile\_linux41.sysload} contains
the parameter \texttt{quiet} to suppress unnecessary messages while
the System Loader is booted. The second parameter references the system
loader configuration file \texttt{sysload.conf} (shown in 
\vref{sub:System-Loader-Configuration})
on a disk of the virtual machine. Note that on the i386 platform usually
the \texttt{block} URI scheme would be used instead of the \texttt{dasd}
URI scheme.

{\scriptsize \verbatiminput{parmfile_linux41.sysload}}


\subsubsection{zipl.conf on System 2}
System 2 is another virtual machine called \texttt{linux40}. This
machine has a kernel and the initial ramdisk of the System Loader
installed locally but gets the boot configuration via the network
from system 1. The initial setup of the network interface is done
via the \texttt{kset=} parameter on the kernel commandline. The 
\texttt{sysload=} parameter references the boot configuration via ftp.

{\scriptsize \verbatiminput{zipl2.conf}}


\subsubsection{System Loader Configuration Used by Both Systems\label{sub:System-Loader-Configuration}\label{sub:sysload.conf-for-s390}}
The following System Loader configuration is used by both systems
and shows various examples for system dependent definitions, device
setup and included configuration parts and how these capabilities
can be combined. 

{\scriptsize \verbatiminput{sysload.conf}}

To show the include mechanism the definition of the generic boot entries
\texttt{reboot}, \texttt{halt} and \texttt{shell} is done via the
following include file:

{\scriptsize \verbatiminput{standard_entries.conf}}

\end{document}
